---
title: "Week 9"
author: "Asger Svenning"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: true
    number_sections: true
toc_title: "Table of Contents"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggforce)
library(patchwork)
library(magrittr)
library(extrafont)
library(ggpubr)

formatScientific <- function(n, latex = F) sapply(n, function(x) {
  
  if (is.na(x)) return(NA)
  if (x == 0) return("0")
  
  negativeFlag <- sign(x) == -1
  
  x <- abs(x)
  log10Base <- log10(x)
  
  if (log10Base < 3 & log10Base > -2) {
    
    out <- if (log10Base < 0) as.character(signif(x, 2)) else as.character(round(x, 2))
    
  } 
  else {
    
    log10Base <- floor(log10Base)
    x <- x * 10^-log10Base
    
    if (round(x, 2) == 1) {
      out <- paste0("10^{", log10Base, "}")
    } 
    else {
      out <- paste0(round(x, 2), " \\times 10^{", log10Base, "}")
    }
  }
  
  out <- stringr::str_remove(out, "^1 ")
  
  if (negativeFlag) out <- paste0("-", out)
  
  if (!latex) return(
    latex2exp::TeX(paste0("$", out, "$"))
  )
  else return(paste0("$", out, "$"))
}, USE.NAMES = F, simplify = T)

theme_set(
  theme_pubr(base_family = "CMU Serif",
             legend = "right") +
    theme(title = element_text(face = "bold",
                               size = 14),
          plot.title = element_text(face = "plain",
                                    hjust = .5,
                                    size = 18),
          legend.title = element_text(hjust = .5))
)


ArchaicSegments <- read_delim("ArchaicSegments.txt", delim = "\t") %>% 
  select(!2)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 6)
```

\newpage

# Total length of Archaic fragments in each individual

```{r}
ArchaicSegments %>% 
  group_by(name) %>% 
  summarize(
    length = sum(length, na.rm = T)
  ) %>% 
  ggplot(aes(name, length)) +
  geom_col(color = "gray85", fill = "gray85") +
  scale_y_continuous(labels = formatScientific) +
  coord_cartesian(expand = F) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "Individual", y = "Length")
```

\newpage

# Total length of Archaic fragments in each population

```{r, fig.height=8}
ArchaicSegments %>% 
  group_by(pop) %>% 
  summarize(
    length = sum(length, na.rm = T)
  ) %>% 
  arrange(pop) %>% 
  mutate(pop = factor(pop)) %>% 
  ggplot(aes(length, pop)) +
  geom_col(color = "gray85", fill = "gray85") +
  scale_x_log10(labels = formatScientific) +
  facet_wrap(~as.numeric(pop)/max(as.numeric(pop)) < 0.5, scales = "free_y") +
  coord_cartesian(expand = F,
                  xlim = c(10^7, 10^10)) +
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = margin(0,2,0,2,"lines")) +
  labs(y = "Population", x = "Length")
```

# Which population has longer fragment reads?

The population "Papuans" from Papua New Guinea has longer reads, probably because they have admixture with both Neanderthals and Denisovans, the latter of which happened more recently.

# What is the length distribution of Archaic fragments for the five different regions

```{r}
ArchaicSegments %>% 
  bind_rows(ArchaicSegments %>% 
              mutate(region = "Combined")) %>% 
  mutate(region = region %>% 
           factor(
             levels = c(
               "CentralAsiaSiberia",
               "EastAsia",
               "Melanesia",
               "SouthAsia",
               "WestEurasia",
               "Combined"
             )
           )) %>% 
  ggplot(aes(length, fill = region)) +
  geom_histogram(aes(y = after_stat(ncount),
                     color = after_scale(fill)),
                 bins = 50,
                 key_glyph = draw_key_point) +
  scale_x_log10(labels = formatScientific) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_brewer(palette = "Dark2") +
  coord_cartesian(expand = F) +
  facet_wrap(~region, ncol = 2) +
  guides(fill = guide_legend(override.aes = list(shape = 21,
                                                 size = 5))) +
  theme(aspect.ratio = 1) +
  labs(x = "Length", y = "Proportion", fill = "Region")
```

\newpage

# What is the average length of fragments for each population and region

```{r, fig.height=8}
meanArchaicLengthByPopulation <- ArchaicSegments %>% 
  arrange(region, pop) %>% 
  mutate(pop = factor(pop, levels = unique(pop))) %>% 
  ggplot(aes(length, pop, color = region)) +
  stat_summary() +
  facet_wrap(~as.numeric(pop)/max(as.numeric(pop)) < 0.5, scales = "free_y") +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(labels = formatScientific,
                     breaks = 10^5 * c(1, 1.3),
                     expand = expansion()) +
  scale_y_discrete(expand = expansion(0, 0.5)) +
  coord_cartesian(xlim = 10^5 * c(0.9, 1.4)) +
  guides(color = guide_legend(nrow = 2)) +
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = margin(0,2,0,2,"lines"),
        legend.position = "bottom") +
  labs(x = "Length", y = "Population", color = "Region")


meanArchaicLengthByRegion <- ArchaicSegments %>% 
  arrange(region) %>% 
  mutate(region = factor(region)) %>% 
  ggplot(aes(length, region, color = region)) +
  stat_summary() +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(labels = formatScientific,
                     breaks = 10^5 * c(1, 1.3),
                     expand = expansion()) +
  scale_y_discrete(expand = expansion(0, .5)) +
  coord_cartesian(xlim = 10^5 * c(0.9, 1.4)) +
  guides(color = guide_legend(nrow = 2)) +
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        plot.margin = margin(0,2,0,2,"lines"),
        legend.position = "bottom") +
  labs(x = "Length", y = "Region", color = "Region")

meanArchaicLengthByPopulation
```

```{r}
meanArchaicLengthByRegion
```


# What can cause different fragment lengths?

Two things can cause different fragment lengths:

1) Time until admixture event
2) Generation time

\newpage

# Mean Archaic segment origin proportion by population and region

```{r, fig.height = 8}
ArchaicSegments %>% 
  select(name, pop, region, start, length, contains("Shared_with")) %>% 
  # mutate(origin = apply(across(contains("Shared_with_")), 1, which.max),
  #        unassigned = rowSums(across(contains("Shared_with_")) > 0) == 0,
  #        origin = c("Altain", "Denisova", "Vindija")[origin],
  #        origin = ifelse(unassigned, "Unassigned", origin))
  pivot_longer(contains("Shared_with"), names_to = "origin",
               names_prefix = "Shared_with_",
               values_to = "snps_origin") %>% 
  group_by(name, pop, region, start, length) %>% 
  summarize(
    unassigned = !any(snps_origin > 0),
    origin = origin[which.max(snps_origin)],
    origin = ifelse(unassigned, "Unassigned", origin),
    .groups = "drop"
  ) %>% 
  group_by(name, pop, region, origin) %>% 
  summarize(
    length = mean(length),
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(pop, region, origin) %>% 
  summarize(
    length = mean(length),
    n = mean(n),
    .groups = "drop"
  ) %>% 
  mutate(
    origin = factor(origin,
                    levels = rev(c("Altai", "Vindija", "Denisova", "Unassigned")))
  ) %>% 
  mutate(region = ifelse(region == "Melanesia", "EastAsia", region)) %>% 
  ggplot(aes(n, pop, fill = origin)) +
  geom_col(aes(color = after_scale(fill))) +
  # scale_x_log10() +
  scale_x_continuous(labels = formatScientific) +
  scale_fill_brewer(palette = "Dark2") +
  coord_cartesian(expand = F) +
  facet_wrap(~region,
             scales = "free") +
  labs(x = "Mean number of segments", y = "Population", fill = "Segment\nOrigin") +
  theme(legend.position = "bottom")
```


```{r}
ArchaicSegments %>% 
  select(name, pop, region, start, length, contains("Shared_with")) %>% 
  # mutate(origin = apply(across(contains("Shared_with_")), 1, which.max),
  #        unassigned = rowSums(across(contains("Shared_with_")) > 0) == 0,
  #        origin = c("Altain", "Denisova", "Vindija")[origin],
  #        origin = ifelse(unassigned, "Unassigned", origin))
  pivot_longer(contains("Shared_with"), names_to = "origin",
               names_prefix = "Shared_with_",
               values_to = "snps_origin") %>% 
  group_by(name, pop, region, start, length) %>% 
  summarize(
    unassigned = !any(snps_origin > 0),
    origin = origin[which.max(snps_origin)],
    origin = ifelse(unassigned, "Unassigned", origin),
    .groups = "drop"
  ) %>% 
  group_by(name, pop, region, origin) %>% 
  summarize(
    length = mean(length),
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(region, origin) %>% 
  summarize(
    length = mean(length),
    n = mean(n),
    .groups = "drop"
  ) %>% 
  mutate(
    origin = factor(origin,
                    levels = rev(c("Altai", "Vindija", "Denisova", "Unassigned")))
  ) %>% 
  mutate(region = ifelse(region == "Melanesia", "EastAsia", region)) %>% 
  ggplot(aes(n, region, fill = origin)) +
  geom_col() +
  scale_x_continuous(labels = formatScientific) +
  scale_fill_brewer(palette = "Dark2") +
  coord_cartesian(expand = F) +
  labs(x = "Mean number of segments", y = "Region", fill = "Segment\nOrigin") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 2, 0, 2, "lines"))
```

# Proportion of Denisova SNP's in individuals closely associated with Denisova's

```{r}
# addIntervals <- function(interval_factor) {
#   intervals <- levels(interval_factor)
#   intervals_num <- intervals %>% 
#     str_extract_all("[0-9\\.\\-]+", simplify = T) %>% 
#     as.numeric %>% 
#     matrix(ncol = 2)
#   
#   interval_width <- intervals_num[, 2] - intervals_num[, 1]
#   interval_width <- round(interval_width, ceiling(abs(log10(mean(interval_width)))) + 1)
#   interval_width <- unique(interval_width)
#   
#   if (length(interval_width) != 1) stop("Cannot add intervals to interval factor with uneven interval widths.")
#   
#   new_intervals <- rbind(
#     intervals_num[1, ] - interval_width,
#     intervals_num,
#     intervals_num[nrow(intervals_num), ] + interval_width) %>% 
#     apply(1, function(x) paste0("(", x[1], ",", x[2], "]"))
#   new_intervals[1] <- new_intervals[1] %>% str_replace("\\(", "[")
#   
#   changed_first_interval <- str_replace(intervals[1], "\\[", "(")
#   
#   interval_factor <- as.character(interval_factor)
#   
#   factor(ifelse(interval_factor == intervals[1], changed_first_interval, interval_factor), levels = new_intervals)
# }

ArchaicSegments %>% 
  select(name, pop, region, snps, contains("Shared_with_")) %>% 
  group_by(name, pop, region) %>% 
  summarize(
    across(contains("Shared_with_"), ~sum(.x)/sum(snps)),
    .groups = "drop"
  ) %>% 
  pivot_longer(contains("Shared_with_"),
               names_to = "origin",
               names_prefix = "Shared_with_",
               values_to = "proportion") %>% 
  filter(origin == "Denisova") %>% 
  mutate(prop_int = cut_width(proportion, 0.005, boundary = 0, dig.lab = 10)) %>% 
  count(region, prop_int, .drop = F) %>% 
  group_by(region) %>% 
  mutate(n = n/sum(n)) %>% 
  ungroup %>% 
  mutate(proportion = str_extract_all(prop_int, "[0-9\\.\\-]+", simplify = T) %>% 
           as.numeric %>% 
           matrix(ncol = 2) %>% 
           rowMeans) %>% 
  ggplot(aes(proportion, n, fill = region)) +
  geom_col(linewidth = 1.25, alpha = 0.75,
           key_glyph = draw_key_point,
           color = "transparent",
           position = position_dodge(preserve = "single")) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) +
  coord_cartesian(expand = F) +
  guides(fill = guide_legend(override.aes = list(size = 5,
                                                  shape = 21))) +
  labs(x = "Proportion of SNP's with Denisova Origin", y = "Proportion of Individuals in Region",
       color = "Region") +
  theme(aspect.ratio = .5)
```

# 





